Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-08-05 manuscripts/reply/reply-main.tex
More editing of reply ms.

However, \citet{Hickerson2013} conclude that the bias is caused by numerical
approximation error due to insufficient computation
(Hypothesis~\ref{hypError}).
They argue the widest of the three priors used by \citet{Oaks2012} would
infrequently produce random samples with many independent population divergence
times as recent as the estimated gene divergence times presented in
\citet{Oaks2012}.
However, this argument assumes the gene divergence times presented in
\citet{Oaks2012} were estimated without error.
These estimates were intended to provide only a rough comparison of the
gene divergence times across the 22 taxa.
These analyses assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations
per generation for all taxa, and are, of course, subject to estimation error.
Furthermore, the branch-length units of the gene trees are in millions of
years, whereas the divergence-time prior of \msb is in generations, thus
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
The argument of \citet{Hickerson2013} that these estimated gene divergence
times should be used to set the upper bound of the uniform prior on divergence
times for the \msb analysis of the same sequence data seems questionable,
especially given our findings presented above regarding the behavior of \msb
when empirically informed uniform priors are employed.

Even if we assume (1) the arbitrary strict clock is correct, (2) gene
divergence times were estimated without error, and (3) all 22 taxa have
one-year generation times, Hickerson et al.'s
(\citeyear{Hickerson2013}) argument only applies to one of the \emph{three}
priors used by \citet{Oaks2012}.
Under these assumptions, the narrowest prior on divergence times used by
\citet{Oaks2012} ($U(0, 5)$) closely mirrors the range of estimates of
gene divergence times (0--5 million years ago).
Applying Hickerson et al.'s (\citeyear{Hickerson2013}) sampling-probability
argument demonstrates this prior is densely populated with samples with large
numbers of divergence parameters with values younger than the estimated gene
divergence estimates.
Thus, if insufficient prior sampling (Hypothesis~\ref{hypError}) is to blame
for the bias, it should be much reduced under the narrow prior on \divt{}.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrowest prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug in
\msb, and after \citet{Oaks2012} corrected the bug, there are many cases where
the narrowest prior performs slightly worse (see panels D--J of Figures 3 and
S12).

To disentangle whether Hypothesis 1 or 2 is the primary cause the method's
erroneous support for simple models, we must look at the different predictions
made by these two phenomena.
For example, numerical error due to insufficient prior sampling
(Hypothesis~\ref{hypError}) should create large variance among posterior
estimates and cause analyses to be highly sensitive to the number of samples
drawn from the prior.
Furthermore, if insufficient prior sampling is \emph{biasing} estimates toward
models with less parameter space, as suggested by \citet{Hickerson2013}, we
expect to see support for these models decrease as sampling from the prior
increases.
\citet{Oaks2012} did not see such sensitivity when they compared prior sample
sizes of $2\e6$, $5\e6$, and $10^7$.

To explore this prediction further, we repeat the analysis of the Philippines
dataset under the intermediate prior used by \citet{Oaks2012} ($\divt{} \sim
U(0, 10)$, $\meanDescendantTheta{} \sim (0.0005, 0.04)$, $\ancestralTheta{}
\sim (0.0005, 0.02)$), using a very large prior sample size of $10^8$.
When we look at the trace of the estimates of the dispersion index of
divergence times (\vmratio{}) as the prior samples accumulate
(Figure~S\ref{figSamplingError}) we do not see the trend predicted by
Hypothesis~\ref{hypError}.
While approximation error is always present for any numerical analysis, it does
not appear to be playing a large role in the biases revealed by the results of
\citet{Oaks2012} or presented above.

A straightforward prediction if strongly weighted marginal likelihoods are
causing the preference for simple models (Hypothesis~\ref{hypML}) is that the
bias should disappear as the model generating the data converges to the prior.
\citet{Oaks2012} tested this prediction by performing 100,000 simulations to
assess the model-choice behavior of \msb when the prior model is correct.
The results confirm the prediction of Hypothesis~\ref{hypML}:
\msb estimates the probability of the one-divergence model quite well (or even
\emph{under}estimates it) when the prior is correct (see Figure 4 of
\citet{Oaks2012}).
We confirmed this same behavior for the model-averaging approach used by
\citet{Hickerson2013} (see SI text and Figure~S\ref{figValidationMCBehavior}).
These results are not clearly predicted if insufficient computation was causing
numerical error.
Even when the prior is correct, due to the discrete uniform prior on the number
of divergence events (\numt{}) implemented in \msb, models with larger numbers
of divergence-time parameters (and thus greater parameter space) will still be
far less densely sampled than those with fewer divergence events
\citep{Oaks2012}.
Thus, the results of the simulations of \citet{Oaks2012} are more consistent
with the fundamental sensitivity of marginal likelihoods to priors
(Hypothesis~\ref{hypML}).

This is further demonstrated by the results presented herein that show the
model-averaging approach of \citet{Hickerson2013} prefers models with narrower
\divt{} priors (Table~\ref{tabModelChoiceEmpirical} and
Figs.~\labelcref{figExclusionSimTau,figExclusionSimProb,figPowerExclusion4})
and fewer \divt{} parameters (Figure~\ref{figPower4}).
In all of these analyses, the same number of random samples were drawn from
each of the prior models.
Thus, while approximation error will always be present in any numerical
approximation method, insufficient prior sampling is an unlikely explanation
for the erroneous support for models with less parameter space.
While analyses that sample each model proportional to their parameter space
could be explored, it is clear that the marginal likelihoods under broad
uniform priors on divergence times will be greater for models with fewer
divergence-time dimensions.

@@

\section{Conclusions}
We demonstrate how the approximate Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that often exclude the true
values of the divergence times the method is trying to estimate.
Our results show that it is difficult to choose an uniformly distributed prior
on divergence times that is broad enough to confidently contain the true values
of parameters while being narrow enough to avoid strongly weighted and
misleading posterior support for models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection, because they can strongly influence the
\emph{true} posterior probabilities of the models we seek to estimate.

>>>>

However, \citet{Hickerson2013} conclude that the bias is caused by numerical
approximation error due to insufficient computation
(Hypothesis~\ref{hypError}).
They argue the widest of the three priors on divergence times used by
\citet{Oaks2012} would infrequently produce random samples with many
independent population divergence times as recent as the estimated gene
divergence times presented in \citet{Oaks2012}.
% They also also suggest that Oaks et al.'s \citeyear{Oaks2012} estimated gene
% divergence times should be used to set the upper bound of the uniform prior on
% divergence times for the \msb analysis of the same sequence data.
% Given the gene-tree estimates were intended to provide only a rough comparison
% of the gene divergence times across the 22 taxa \citep{Oaks2012}, and our
% findings presented above regarding the behavior of \msb when empirically
% informed uniform priors are employed, such an approach should not be used.
However, this sampling-probability argument is based on some questionable
assumptions.
Oaks et al.'s (\citeyear{Oaks2012}) gene-tree estimates were intended to
provide only a rough comparison of the gene divergence times across the 22 taxa
and assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations per
generation for all taxa.
Furthermore, because the branch-length units of the gene trees are in millions
of years whereas the divergence-time prior of \msb is in generations,
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
More importantly, even if we assume (1) the arbitrary strict clock is correct,
(2) gene divergence times were estimated without error, and (3) all 22 taxa
have one-year generation times, Hickerson et al.'s (\citeyear{Hickerson2013})
argument actually demonstrates that the models used by \citet{Oaks2012} with
narrower priors on divergence times are densely populated with samples with
large numbers of divergence parameters with values younger than the estimated
gene divergence estimates.
Thus, if \citet{Hickerson2013} are correct, analyses under these narrow priors
should be much less biased toward clustered models of divergence.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrowest prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug in
\msb, and after \citet{Oaks2012} corrected the bug, there are many cases where
the narrowest prior performs slightly worse (see panels D--J of Figures 3 and
S12).

To disentangle whether Hypothesis 1 or 2 is the primary cause the method's
erroneous support for simple models, we must look at the different predictions
made by these two phenomena.
For example, numerical error due to insufficient prior sampling
(Hypothesis~\ref{hypError}) should create large variance among posterior
estimates and cause analyses to be highly sensitive to the number of samples
drawn from the prior.
Furthermore, if insufficient prior sampling is \emph{biasing} estimates toward
models with less parameter space we expect to see support for these models
decrease as sampling from the prior increases.
\citet{Oaks2012} did not see such sensitivity when they compared prior sample
sizes of $2\e6$, $5\e6$, and $10^7$.

To explore this prediction further, we repeat the analysis of the Philippines
dataset under the intermediate prior used by \citet{Oaks2012} ($\divt{} \sim
U(0, 10)$, $\meanDescendantTheta{} \sim (0.0005, 0.04)$, $\ancestralTheta{}
\sim (0.0005, 0.02)$), using a very large prior sample size of $10^8$.
When we look at the trace of the estimates of the dispersion index of
divergence times (\vmratio{}) as the prior samples accumulate
(Figure~S\ref{figSamplingError}) we do not see the trend predicted by
Hypothesis~\ref{hypError}.
While approximation error is always present in any numerical analysis, it does
not appear to be playing a large role in the biases revealed by the results of
\citet{Oaks2012} or presented above.

A straightforward prediction if strongly weighted marginal likelihoods are
causing the preference for simple models (Hypothesis~\ref{hypML}) is that the
bias should disappear as the model generating the data converges to the prior.
\citet{Oaks2012} tested this prediction by performing 100,000 simulations to
assess the model-choice behavior of \msb when the prior model is correct.
The results confirm the prediction of Hypothesis~\ref{hypML}:
\msb estimates the probability of the one-divergence model quite well (or even
\emph{under}estimates it) when the prior is correct (see Figure 4 of
\citet{Oaks2012}).
We confirmed this same behavior for the model-averaging approach used by
\citet{Hickerson2013} (see SI text and Figure~S\ref{figValidationMCBehavior}).
These results are not clearly predicted if insufficient computation was causing
numerical error (Hypothesis~\ref{hypError}).
Even when the prior is correct, due to the discrete uniform prior on the number
of divergence events (\numt{}) implemented in \msb, models with larger numbers
of divergence-time parameters (and thus greater parameter space) will still be
far less densely sampled than those with fewer divergence events
\citep{Oaks2012}.
Thus, the results of the simulations of \citet{Oaks2012} are more consistent
with the fundamental sensitivity of marginal likelihoods to priors
(Hypothesis~\ref{hypML}).

This is further demonstrated by the results presented herein that show the
model-averaging approach of \citet{Hickerson2013} prefers models with narrower
\divt{} priors (Table~\ref{tabModelChoiceEmpirical} and
Figs.~\labelcref{figExclusionSimTau,figExclusionSimProb,figPowerExclusion4})
and fewer \divt{} parameters (Figure~\ref{figPower4}).
For these model-averaging analyses, insufficient prior sampling
(Hypothesis~\ref{hypError}) is an unlikely explanation for the erroneous support
for models with less parameter space, because (1) all of the prior models share
the same dimensionality, and (2) the same number of random samples were drawn
from each of the prior models.
These results are predicted by Hypothesis~\ref{hypML}, because the marginal
likelihoods will be higher for models with narrower priors on divergence times
and fewer divergence-time dimensions (these models integrate over less space
with large prior weight and small likelihood).

@@

\section{Conclusions}
We demonstrate how the approximate Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that often exclude the true
values of the divergence times.
Our results show that it is difficult to choose an uniformly distributed prior
on divergence times that is broad enough to confidently contain the true values
of parameters while being narrow enough to avoid strongly weighted and
misleading posterior support for models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection, because they can strongly influence the
\emph{true} posterior probabilities of the models we seek to estimate.
